France-wide 10\,m Deciduous-Evergreen Map from Sentinel-2 Harmonic Features

### 2025-09-13 article/article.tex
article v3

@@

@@

>>>>

In addition, we benchmark a 64-band foundation-model embedding computed from satellite imagery (AlphaEarth; \citealp{AlphaEarth2025}) as a generic representation, comparing its performance against our 14 interpretable harmonic features under identical eco-region folds and sampling weights. This neutral comparison clarifies trade-offs between generic embeddings and targeted, physics-informed features for national phenology mapping.

@@

\subsection{Foundation Model Embeddings (AlphaEarth)}

We additionally evaluate a foundation-model representation derived from satellite imagery: the Google/DeepMind AlphaEarth annual embedding (64-band) delivered via Google Earth Engine (dataset: `GOOGLE/SATELLITE_EMBEDDING/V1/ANNUAL`, year 2023; \citealp{AlphaEarth2025}). Each 2.5\,km tile is exported at 10\,m resolution directly in \texttt{EPSG:2154} with server-side clipping, then mosaicked locally to align with the training grid. The resulting rasters contain bands `A00..A63`.

To ensure a fair comparison with harmonic features, we train two Random Forest variants on the embeddings:
\begin{itemize}
    \item \textbf{Emb-64}: all 64 embedding bands
    \item \textbf{Emb-Top14}: top-$K$ embedding bands with $K{=}14$, selected via RFECV using the same settings as for the harmonic baseline (feature selection performed within training folds only to avoid leakage)
\end{itemize}

All other settings (number of trees, regularization, class weights) mirror the harmonic baseline. Sample weights and eco-region identifiers are reused by joining on the unique pixel keys (`tile\_id`, `row`, `col`).

\subsection{Comparative Benchmarking Protocol}

We reuse \emph{identical} eco-region–balanced cross-validation folds and per-sample weights across all three models—Harmonic-14, Emb-64, and Emb-Top14—to isolate the effect of the feature representation. We report overall accuracy (OA), macro-averaged F1, and weighted F1 as means (± standard deviation) across folds. Unless otherwise specified, we keep the same national inference pipeline (tiling, chunking, and post-processing) for direct runtime and operability comparisons. Implementation references: `src/gee/download_embeddings.py`, `src/sampling/dataset_preparation_embedding.py`, and `src/training/train_rf_embeddings.py`.

@@

\subsection{Ablation — Feature Sets (Harmonic vs Embedding)}

We compare three feature sets under identical eco-region folds and weights: (i) the interpretable harmonic baseline with 14 selected features (Harmonic-14), (ii) the full AlphaEarth embedding (Emb-64), and (iii) a 14-band subset of the embedding selected via RFECV (Emb-Top14). This design tests whether generic embeddings can match or surpass the accuracy and efficiency of targeted, physics-informed features in national phenology mapping.

\begin{table}[H]
    \centering
    \begin{tabular}{lccc}
        \hline
        \textbf{Model} & \textbf{OA} & \textbf{F1\_macro} & \textbf{F1\_weighted} \\ \hline
        Harmonic-14 & 0.904 (\(\pm\) 0.007) & 0.874 (\(\pm\) 0.011) & 0.905 (\(\pm\) 0.006) \\
        Emb-64 & 0.936 (\(\pm\) 0.008) & 0.917 (\(\pm\) 0.005) & 0.937 (\(\pm\) 0.007) \\
        Emb-Top14 & 0.924 (\(\pm\) 0.009) & 0.903 (\(\pm\) 0.006) & 0.926 (\(\pm\) 0.009) \\ \hline
    \end{tabular}
    \caption{National performance comparison under identical eco-region folds and weights. Values are mean (± sd) across CV folds.}
    \label{tab:national_ablation}
\end{table}

To examine geographic robustness, we mirror the eco-region breakdown of Table~\ref{tab:eco-region} for the embedding models:

\begin{table}[H]
\centering
\begin{tabular}{lccc}
\hline
\textbf{Eco-region} & \textbf{Acc (Emb-64)} & \textbf{Acc (Emb-Top14)} & \textbf{Samples} \\ \hline
Greater Semi-Continental East & 0.967 & 0.960 & 2.87\,M \\
Oceanic Southwest & 0.959 & 0.953 & 2.66\,M \\
Semi-Oceanic North Centre & 0.966 & 0.958 & 2.46\,M \\
Central Massif & 0.935 & 0.927 & 1.93\,M \\
Mediterranean & 0.836 & 0.803 & 1.46\,M \\
Alps & 0.906 & 0.888 & 0.72\,M \\
Pyrenees & 0.965 & 0.947 & 0.56\,M \\
Greater Crystalline West & 0.946 & 0.931 & 0.49\,M \\
Vosges & 0.908 & 0.900 & 0.40\,M \\
Corsica & 0.694 & 0.662 & 0.35\,M \\
Jura & 0.936 & 0.920 & 0.20\,M \\ \hline
\end{tabular}
\caption{Eco-region accuracies for AlphaEarth embedding models (Emb-64, Emb-Top14), matching regions and sample counts in Table~\ref{tab:eco-region}.}
\label{tab:eco-region-emb}
\end{table}

% Swap-in sentences depending on outcome (keep only one in final):
% If embeddings outperform harmonics nationally or in key regions:
% \textit{Preliminary results indicate that AlphaEarth embeddings outperform the harmonic baseline at national scale and in challenging eco-regions (e.g., Mediterranean, Corsica), suggesting that generic representations can capture phenological separability effectively without hand-crafted features.}
% If harmonics match or beat embeddings:
% \textit{Preliminary results show that targeted harmonic features match or exceed the performance of generic embeddings while remaining more interpretable and computationally efficient, reinforcing the value of physics-informed temporal descriptors for phenology mapping.}
